Pre working group data exploration: Mainly wrt coral cover

### 2024-07-10 julianna_data_exploration.Rmd
Started playing around with SACs

# load packages
require(here) # relative file paths
require(tidyverse) # data wrangling
require(vegan) # community/diversity analyses
require(lme4) # glmer()
require(emmeans) # emmeans()
require(viridis) # color palette
require(MoMAColors) # another color palette
```

@@

```{r}
# matrix of just species and their relative percent covers in the same order as the algae tibble
algae %>% 
  select(-year, -location, -site, -island_side, -habitat, -site_habitat, -transect, -quadrat) -> sppMat
```

@@

algSum %>% 
  full_join(allbenthos) %>% 
  ggplot(aes(x = coral, y = AlgSum, color = year)) +
  geom_point() +
  facet_grid(site~habitat)

@@

```{r}
coralAlgDiv %>% 
  ggplot(aes(x = coral, y = Richness, color = year, group = Richness)) +
  geom_jitter() + 
  facet_grid(habitat~site) +
  theme_classic()
```

@@

@@

```{r make a df for island side/habitat}
macroalgae %>% 
  mutate(macroalgae_sum = rowSums(macroalgae[startMAlg:endMAlg])) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side, macroalgae_sum) %>% 
  full_join(coralAlgDiv) %>% 
  full_join(algae) %>% 
  rowwise() %>% 
  mutate(turf_sum = sum(algal_turf, damselfish_turf),
         CCA_sum = sum(lithophyllum_kotschyanum, crustose_corallines, neogoniolithon_brassica_florida, sporolithon_sp)) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side, turf_sum, CCA_sum, coral, macroalgae_sum) %>% 
  # pivot key columns longer
  pivot_longer(cols = c(turf_sum, CCA_sum, coral, macroalgae_sum),
               names_to = "substrate",
               values_to = "sum_cover") %>% 
  # summarize and group quadrats by island/habitat
  group_by(year, habitat, island_side, substrate) %>% 
  summarise(mean_cov = mean(sum_cover),
            sd_cov = sd(sum_cover, na.rm = TRUE),
            n = n(),
            CI95 = 1.96*(sd_cov)/sqrt(n)) %>% 
  # relevel for graphing
  mutate(island_side = factor(island_side, levels = c("north", "east", "west")),
         habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) -> quadHabIS

@@

@@

transectHab %>% 
  ggplot(aes(x = year, y = mean_cov, group = substrate, fill = substrate)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
  scale_fill_manual(values = c(
    moma.colors("Klein", type = "discrete")[2], # CCA
    moma.colors("Klein", type = "discrete")[1], # coral
    moma.colors("Klein", type = "discrete")[5], # macro
    moma.colors("Klein", type = "discrete")[4] # turf
  )) +
  facet_wrap(~habitat) +
  ylab("Mean cover (%)") +
  xlab("Year")
```

@@

@@

## Distance matrix

@@

```{r}
sppMat %>% 
  mutate(AlgSum = rowSums(across(where(is.numeric)))) %>% 
  filter(AlgSum == 0) %>% dim() # 672 with no algae at all
```

@@

```{r}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

@@

@@

```{r}
brayMat <- vegdist(sppMat0, "bray") # 1 is maximum dissimilarity
```

@@

```{r}
habBetaDisper <- vegan::betadisper(brayMat, sppMeta0$hab_year)

# plot of dispersion test
plot(habBetaDisper, sub = "") # lol
boxplot(habBetaDisper, main = "Habitat", xlab = "") 
```

>>>>

# load packages
require(here) # relative file paths
require(tidyverse) # data wrangling
require(vegan) # community/diversity analyses
require(lme4) # glmer()
require(emmeans) # emmeans()
require(viridis) # color palette
require(MoMAColors) # another color palette
require(patchwork) # putting figures together
require(BiodiversityR)
require(ggsci) # scientific jounral and sci-fi themed color palette
```

@@

```{r}
# matrix of just species and their relative percent covers in the same order as the algae tibble
algae %>% 
  select(-year, -location, -site, -island_side, -habitat, -site_habitat, -transect, -quadrat) -> sppMat

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, "_", hab_year)) -> sppMeta
```

@@

algSum %>% 
  full_join(allbenthos) %>% 
  ggplot(aes(x = coral, y = AlgSum, color = year)) +
  geom_point() +
  facet_grid(site~habitat) +
  ylab("Algal cover (%)") +
  xlab("Coral cover (%)")

@@

```{r}
coralAlgDiv %>% 
  ggplot(aes(x = coral, y = Richness, color = year, group = Richness)) +
  geom_jitter() + 
  facet_grid(habitat~island_side) +
  theme_classic()
```

As boxplots

```{r}
coralAlgDiv %>% 
  ggplot(aes(x = Richness, y = coral, color = island_side, group = Richness)) +
  geom_boxplot() + 
  facet_grid(habitat~island_side) +
  scale_color_manual(values = c("darkblue", "darkred", "darkgreen")) +
  theme_classic()

@@

# Species accummulation curves

## Set special theme

From https://rpubs.com/Roeland-KINDT/694021

```{r}
BioR.theme <- theme(
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid = element_blank(),
        axis.line = element_line("gray25"),
        text = element_text(size = 12),
        axis.text = element_text(size = 10, colour = "gray25"),
        axis.title = element_text(size = 14, colour = "gray25"),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 14),
        legend.key = element_blank())
```

## By habitat (all time)

Noam suggested doing this on July 10, 2024 in response to the richness plots. To learn how to do this, I used this tutorial: https://rpubs.com/Roeland-KINDT/694021

```{r accumpcomp}
# use accumcomp() to calculate accumulated species richness
# conditioned = FALSE switches estimation of standard deviation to a formula based on the estimation of extrapolated species richness
# returns data on species accumulations separately for each different level of a factor
accum.Hab <- accumcomp(sppMat, # spp matrix
                     y = sppMeta, # metadata df (same length)
                     factor = 'habitat', # factor to group by in the metadata file
    method='exact', conditioned = FALSE, plotit = FALSE)
```

To plot, use accumcomp.long() to get it into long/ggplot format

```{r make long}
accum.Hablong <- accumcomp.long(accum.Hab, ci=NA, 
                              label.freq=2000) # lebel.freq adjusts the labelit variabel to show how often a symbol will be plotted on the SAC
```

```{r plot}
accum.Hablong %>% 
  mutate(Grouping = factor(Grouping, levels = c("fringing", "backreef", "outer_10", "outer_17"))) %>% 
  ggplot(aes(x = Sites, y = Richness, ymax = UPR, ymin = LWR)) +
  scale_x_continuous(expand=c(0, 1), sec.axis = dup_axis(labels=NULL, name=NULL)) + # this moves it so zero is right at the origin
  scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) + # this makes it more like a box
  geom_line(aes(colour=Grouping), linewidth=2) + # adds line
  #geom_point(data=subset(accum.Hablong, labelit==TRUE), 
  #             aes(colour=Grouping, shape=Grouping), size=.5) +
  geom_ribbon(aes(colour=Grouping), alpha=0.2, show.legend=FALSE) + 
  BioR.theme +
  scale_colour_viridis_d(direction = -1) +
  labs(x = "Quadrats", y = "Richness", colour = "Haitat", shape = "Haitat")
```

## By island side (all time)

```{r accumpcomp}
# use accumcomp() to calculate accumulated species richness
# conditioned = FALSE switches estimation of standard deviation to a formula based on the estimation of extrapolated species richness
# returns data on species accumulations separately for each different level of a factor
accum.IS <- accumcomp(sppMat, # spp matrix
                     y = sppMeta, # metadata df (same length)
                     factor = 'island_side', # factor to group by in the metadata file
    method='exact', conditioned = FALSE, plotit = FALSE)
```

To plot, use accumcomp.long() to get it into long/ggplot format

```{r make long}
accum.ISlong <- accumcomp.long(accum.IS, ci=NA, 
                              label.freq=2000) # lebel.freq adjusts the labelit variabel to show how often a symbol will be plotted on the SAC
```

```{r plot}
accum.ISlong %>% 
  mutate(Grouping = factor(Grouping, levels = c("north", "east", "west"))) %>% 
  ggplot(aes(x = Sites, y = Richness, ymax = UPR, ymin = LWR)) +
  scale_x_continuous(expand=c(0, 1), sec.axis = dup_axis(labels=NULL, name=NULL)) + # this moves it so zero is right at the origin
  scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) + # this makes it more like a box
  geom_line(aes(colour=Grouping), linewidth=2) + # adds line
  #geom_point(data=subset(accum.Hablong, labelit==TRUE), 
  #             aes(colour=Grouping, shape=Grouping), size=.5) +
  geom_ribbon(aes(colour=Grouping), alpha=0.2, show.legend=FALSE) + 
  BioR.theme +
  scale_color_manual(values = c("darkblue", "darkred", "darkgreen")) +
  labs(x = "Quadrats", y = "Richness", colour = "Haitat", shape = "Haitat")
```

## By habitat-time

Noam suggested doing this on July 10, 2024 in response to the richness plots. To learn how to do this, I used this tutorial: https://rpubs.com/Roeland-KINDT/694021

```{r accumpcomp}
# use accumcomp() to calculate accumulated species richness
# conditioned = FALSE switches estimation of standard deviation to a formula based on the estimation of extrapolated species richness
# returns data on species accumulations separately for each different level of a factor
accum.HabYr <- accumcomp(sppMat, # spp matrix
                     y = sppMeta, # metadata df (same length)
                     factor = 'hab_year', # factor to group by in the metadata file
    method='exact', conditioned = FALSE, plotit = FALSE)
```

To plot, use accumcomp.long() to get it into long/ggplot format

```{r make long}
accum.HabYrlong <- accumcomp.long(accum.HabYr, ci=NA, 
                              label.freq=2000) %>% # lebel.freq adjusts the labelit variabel to show how often a symbol will be plotted on the SAC
  # get metadata back
  rowwise() %>% 
  mutate(Year = (str_split(Grouping, pattern = "_")[[1]])[length(str_split(Grouping, pattern = "_")[[1]])]) %>% 
  mutate(Habitat = case_when(
    str_split(Grouping, pattern = "_")[[1]][[2]] == "17" ~ "outer_17",
    str_split(Grouping, pattern = "_")[[1]][[2]] == "10" ~ "outer_10",
    str_split(Grouping, pattern = "_")[[1]][[1]] == "fringing" ~ "fringing",
    str_split(Grouping, pattern = "_")[[1]][[1]] == "backreef" ~ "backreef",
    TRUE ~ "Error"
    )) %>% 
  mutate(Habitat = factor(Habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) 
```

```{r plot}
accum.HabYrlong %>% 
  ggplot(aes(x = Sites, y = Richness, ymax = UPR, ymin = LWR)) +
  scale_x_continuous(expand=c(0, 1), sec.axis = dup_axis(labels=NULL, name=NULL)) + # this moves it so zero is right at the origin
  scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) + # this makes it more like a box
  geom_line(aes(colour=Year), linewidth=2) + # adds line
  #geom_point(data=subset(accum.Hablong, labelit==TRUE), 
  #             aes(colour=Grouping, shape=Grouping), size=.5) +
  #geom_ribbon(aes(colour=Year), alpha=0.2, show.legend=FALSE) + 
  BioR.theme +
  scale_colour_viridis_d(direction = -1) +
  facet_wrap(~Habitat) +
  labs(x = "Quadrats", y = "Richness", colour = "Haitat", shape = "Haitat")
```

## By habitat-time-island side

```{r accumpcomp}
# use accumcomp() to calculate accumulated species richness
# conditioned = FALSE switches estimation of standard deviation to a formula based on the estimation of extrapolated species richness
# returns data on species accumulations separately for each different level of a factor
accum.HabISYr <- accumcomp(sppMat, # spp matrix
                     y = sppMeta, # metadata df (same length)
                     factor = 'island_side_hab_year', # factor to group by in the metadata file
    method='exact', conditioned = FALSE, plotit = FALSE)
```

To plot, use accumcomp.long() to get it into long/ggplot format

```{r make long}
accum.HabISYrlong <- accumcomp.long(accum.HabISYr, ci=NA, 
                              label.freq=2000) %>% # lebel.freq adjusts the labelit variabel to show how often a symbol will be plotted on the SAC
  # get metadata back
  rowwise() %>% 
  mutate(Year = (str_split(Grouping, pattern = "_")[[1]])[length(str_split(Grouping, pattern = "_")[[1]])]) %>% 
  mutate(Habitat = case_when(
    str_split(Grouping, pattern = "_")[[1]][[3]] == "17" ~ "outer_17",
    str_split(Grouping, pattern = "_")[[1]][[3]] == "10" ~ "outer_10",
    str_split(Grouping, pattern = "_")[[1]][[2]] == "fringing" ~ "fringing",
    str_split(Grouping, pattern = "_")[[1]][[2]] == "backreef" ~ "backreef",
    TRUE ~ "Error"
    )) %>% 
  mutate(island_side = str_split(Grouping, pattern = "_")[[1]][[1]]) %>% 
  mutate(Habitat = factor(Habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) 
```

```{r plot}
accum.HabISYrlong %>% 
  ggplot(aes(x = Sites, y = Richness, ymax = UPR, ymin = LWR)) +
  scale_x_continuous(expand=c(0, 1), sec.axis = dup_axis(labels=NULL, name=NULL)) + # this moves it so zero is right at the origin
  scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) + # this makes it more like a box
  geom_line(aes(colour=Year), linewidth=1) + # adds line
  #geom_point(data=subset(accum.Hablong, labelit==TRUE), 
  #             aes(colour=Grouping, shape=Grouping), size=.5) +
  #geom_ribbon(aes(colour=Year), alpha=0.2, show.legend=FALSE) + 
  BioR.theme +
  scale_colour_viridis_d(direction = -1) +
  facet_grid(island_side~Habitat) +
  labs(x = "Quadrats", y = "Richness")
```

@@

```{r make a df for island side/habitat}
macroalgae %>% 
  mutate(macroalgae_sum = rowSums(macroalgae[startMAlg:endMAlg])) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side, macroalgae_sum) %>% 
  full_join(coralAlgDiv) %>% 
  full_join(algae) %>% 
  rowwise() %>% 
  mutate(turf_sum = sum(algal_turf, damselfish_turf),
         CCA_sum = sum(lithophyllum_kotschyanum, crustose_corallines, neogoniolithon_brassica_florida, sporolithon_sp)) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side, turf_sum, CCA_sum, coral, macroalgae_sum) %>% 
  # pivot key columns longer
  pivot_longer(cols = c(turf_sum, CCA_sum, coral, macroalgae_sum),
               names_to = "substrate",
               values_to = "sum_cover") -> quad4sum

quad4sum %>% 
  # summarize and group quadrats by island/habitat
  group_by(year, habitat, island_side, substrate) %>% 
  summarise(mean_cov = mean(sum_cover),
            sd_cov = sd(sum_cover, na.rm = TRUE),
            n = n(),
            CI95 = 1.96*(sd_cov)/sqrt(n)) %>% 
  # relevel for graphing
  mutate(island_side = factor(island_side, levels = c("north", "east", "west")),
         habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) -> quadHabIS

@@

## Quadrat-level habitat

```{r make a df for island side/habitat}
quad4sum %>% 
  # summarize and group quadrats by island/habitat
  group_by(year, habitat, substrate) %>% 
  summarise(mean_cov = mean(sum_cover),
            sd_cov = sd(sum_cover, na.rm = TRUE),
            n = n(),
            CI95 = 1.96*(sd_cov)/sqrt(n)) %>% 
  # relevel for graphing
  mutate(habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) %>% 
  ggplot(aes(x = year, y = mean_cov, group = substrate, fill = substrate)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
  scale_fill_manual(values = c(
    moma.colors("Klein", type = "discrete")[2], # CCA
    moma.colors("Klein", type = "discrete")[1], # coral
    moma.colors("Klein", type = "discrete")[5], # macro
    moma.colors("Klein", type = "discrete")[4] # turf
  )) +
  facet_wrap(~habitat, nrow = 1) +
  ylab("Mean cover (%)") +
  xlab("Year")

@@

transectHab %>% 
  ggplot(aes(x = year, y = mean_cov, group = substrate, fill = substrate)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
  scale_fill_manual(values = c(
    moma.colors("Klein", type = "discrete")[2], # CCA
    moma.colors("Klein", type = "discrete")[1], # coral
    moma.colors("Klein", type = "discrete")[5], # macro
    moma.colors("Klein", type = "discrete")[4] # turf
  )) +
  facet_wrap(~habitat, nrow = 1) +
  ylab("Mean cover (%)") +
  xlab("Year")
```

@@

## Quadrat-level island side

```{r}
quad4sum %>% 
  # summarize and group quadrats by island/habitat
  group_by(year, island_side, substrate) %>% 
  summarise(mean_cov = mean(sum_cover),
            sd_cov = sd(sum_cover, na.rm = TRUE),
            n = n(),
            CI95 = 1.96*(sd_cov)/sqrt(n)) %>% 
  # relevel for graphing
  mutate(island_side = factor(island_side, levels = c("north", "east", "west"))) %>% 
  ggplot(aes(x = year, y = mean_cov, group = substrate, fill = substrate)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
  scale_fill_manual(values = c(
    moma.colors("Klein", type = "discrete")[2], # CCA
    moma.colors("Klein", type = "discrete")[1], # coral
    moma.colors("Klein", type = "discrete")[5], # macro
    moma.colors("Klein", type = "discrete")[4] # turf
  )) +
  facet_wrap(~island_side, nrow = 1) +
  ylab("Mean cover (%)") +
  xlab("Year")
```

## Plot by substrate with habitat as the facet

By transect

```{r plot}
transectHabIS %>% 
  ggplot(aes(x = year, y = mean_cov, group = habitat, fill = habitat)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
  scale_fill_viridis_d(direction = -1) +
  facet_grid(substrate ~ island_side) +
  ylab("Mean cover (%)") +
  xlab("Year")
```

By quadrat

```{r}
quad4sum %>% 
  # summarize and group quadrats by island/habitat
  group_by(year, habitat, island_side, substrate) %>% 
  summarise(mean_cov = mean(sum_cover),
            sd_cov = sd(sum_cover, na.rm = TRUE),
            n = n(),
            CI95 = 1.96*(sd_cov)/sqrt(n)) %>% 
  # relevel for graphing
  mutate(habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17")),
         island_side = factor(island_side, levels = c("north", "east", "west"))) %>% 
  ggplot(aes(x = year, y = mean_cov, group = habitat, fill = habitat)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
  scale_fill_viridis_d(direction = -1) +
  facet_grid(substrate ~ island_side) +
  ylab("Mean cover (%)") +
  xlab("Year")
```

@@

## Distance matrix: habitat-year

@@

```{r look for zeros}
sppMat %>% 
  mutate(AlgSum = rowSums(across(where(is.numeric)))) %>% 
  filter(AlgSum == 0) %>% dim() # 672 with no algae at all
```

@@

```{r get spp mat df}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

@@

Start with just one habitat at a time

### Fringing

```{r get fringe mat}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

# spp matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>% 
  filter(habitat == "fringing") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgSum) -> sppMat0fring

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "fringing") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year)) -> sppMeta0fring
```

@@

```{r calculate distances}
brayMatFring <- vegdist(sppMat0fring, "bray") # 1 is maximum dissimilarity
```

@@

```{r calculate beta dispersion}
yrBetaDisperFring <- vegan::betadisper(brayMatFring, sppMeta0fring$year) # warning: some squared distances are negative and changed to zero (ask Jamie??)

# plot of dispersion test
plot(yrBetaDisperFring, sub = "") 
boxplot(yrBetaDisperFring, main = "Fringing reef", xlab = "") 
```

### Backreef

```{r get backreef df}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

# spp matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>% 
  filter(habitat == "backreef") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgSum) -> sppMat0back

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "backreef") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year)) -> sppMeta0back
```

```{r calculate distance}
brayMatBack <- vegdist(sppMat0back, "bray") # 1 is maximum dissimilarity
```

Calculate beta disperion

```{r calculate beta dispersion}
yrBetaDisperBack <- vegan::betadisper(brayMatBack, sppMeta0back$year)

# plot of dispersion test
plot(yrBetaDisperBack, sub = "") # lol
boxplot(yrBetaDisperBack, main = "Backreef", xlab = "") 
```

### Outer 10

```{r get outer 10 df}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

# spp matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>% 
  filter(habitat == "outer_10") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgSum) -> sppMat0outer10

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "outer_10") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year)) -> sppMeta0outer10
```

```{r calculate distance}
brayMatOuter10 <- vegdist(sppMat0outer10, "bray") # 1 is maximum dissimilarity
```

Calculate beta disperion

```{r calculate beta dispersion}
yrBetaDisperOuter10 <- vegan::betadisper(brayMatOuter10, sppMeta0outer10$year) # error: some squared distances are negative and changed to zero

# plot of dispersion test
plot(yrBetaDisperOuter10, sub = "") # lol
boxplot(yrBetaDisperOuter10, main = "Outer 10", xlab = "") 
```

### Outer 17

```{r get outer 17 df}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

# spp matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>% 
  filter(habitat == "outer_17") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgSum) -> sppMat0outer17

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "outer_17") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year)) -> sppMeta0outer17
```

```{r calculate distance}
brayMatOuter17 <- vegdist(sppMat0outer17, "bray") # 1 is maximum dissimilarity
```

Calculate beta disperion

```{r calculate beta dispersion}
yrBetaDisperOuter17 <- vegan::betadisper(brayMatOuter17, sppMeta0outer17$year) # error: some squared distances are negative and changed to zero

# plot of dispersion test
plot(yrBetaDisperOuter17, sub = "") # lol
boxplot(yrBetaDisperOuter17, main = "Outer 17", xlab = "") 
```

## Plot together

How??

```{r}
```

### Save dispersions

```{r}
# save(yrBetaDisperFring, file = here("data/fringingyrdispersion.Rdata"))
# save(yrBetaDisperBack, file = here("data/backyrdispersion.Rdata"))
# save(yrBetaDisperOuter10, file = here("data/outer10yrdispersion.Rdata"))
# save(yrBetaDisperOuter17, file = here("data/outer17yrdispersion.Rdata"))
```

### Load dispersions

```{r}
load(here("data/fringingyrdispersion.Rdata"))
load(here("data/backyrdispersion.Rdata"))
load(here("data/outer10yrdispersion.Rdata"))
load(here("data/outer17yrdispersion.Rdata"))
```
